FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2017-10-18 NewVersion/sections/Formalization.tex
Fix.

\paragraph{Methods}
Original methods and hierarchically overriding methods share the same syntax in our model for simplicity.
The concrete method declaration $\method{I}{m}{I_x}{x}{J}{e}$ introduces a
method named $m$ with result type $I$, parameters $\overline{x}$ of
type $\overline{I_x}$ and the overriding target $J$. The body of the
method simply includes the returned expression $e$. Notably, we have introduced the
\kwoverride{} keyword to override an old implementation of the
method. If the interface that it overrides is exactly the enclosing
interface, then such a method is seen as originally defined; otherwise it is an overriding method. The definition
of abstract methods is written as $\absmethod{I}{m}{I_x}{x}{J}$, which is
similar to a concrete method but without the method body. 
For simplicity, overloading is not modelled for methods, which
implies that we can uniquely identify a method by its name.

\paragraph{Expressions \& Values}
Expressions can be standard constructs such as variables, method
invocation, object creation, together with cast expressions. 
Object creation is represented by $\new I$. Fields and primitive types are not modelled in \MIM{}. 
The casts are merely safe up-casts, and in fact, they can be viewed as
annotated expressions, where the annotation indicates its static type.
The coexistence of static and dynamic types is the key to hierarchical dispatch.
A value
``$(I)\new{J}$''
is the final result of multiple reduction steps evaluating an
expression.

@@

\textsc{(T-Invk)} is the typing rule for method invocation.
Naturally, the receiver and the arguments are required to be well-typed.
$\mbody$ is our key function for method lookup that implements the
hierarchical dispatching algorithm. 
The hierarchical invocation $\mbody(m, I, J)$ means ``finding most specific method $m$
above $I$ and along path $J$'', and usually $I$ denotes the dynamic type while $J$
denotes the static type. 
The formal definition of $\mbody(m, I, J)$ will be introduced in
Section~\ref{sec:auxdefs}.
In \textsc{(T-Invk)}, as the compilation should not be aware
of the dynamic type, it only requires that invoking $m$ is valid for the static type of the
receiver. The result of $\mbody$ contains the interface that provides the most specific implementation,
the parameters and the return type. We use underscore for the return expression, to imply that an empty return expression
from an abstract method is also acceptable.
Note that we accept implicit upcasts on the arguments.
% \textsc{(T-PathInvk)} is the typing judgement for a path invocation. Besides the conditions of \textsc{(T-Invk)}, \textsc{(T-PathInvk)} requires the type of receiver to be the subtype of the specified path type. 
% and \textsc{(T-SuperInvk)} checks if the enclosing type directly extends the specified super type.

@@

\textsc{(T-Method)} is more interesting since a method can either be an original method or a hierarchical overriding, though
they share the same syntax and method typing rule. $\mostSpecific(m, I, J)$ is a fundamental function,
used to find ``the most specific interfaces that are above $I$ and
along path $J$, and originally defines $m$'' (see
Section~\ref{sec:auxdefs} for full definition).
By ``most specific interfaces'',
it implies that the inherited super types are excluded. Thus the condition $\mostSpecific(m, I, J) = \{J\}$ indicates a characteristic of a hierarchical overriding: it must override an original method; the overriding is direct and there does not exist any other original method $m$ in between.
Then $\mbody(m, J, J)$ provides the type of the original method, so hierarchical overriding has to preserve the type. Finally the return expression
needs to be type-checked, and there is also an implicit cast to the declared return type. For the definition of an original method, $I$ equals $J$ and the rule is straightforward. \textsc{(T-AbsMethod)} is a similar rule but works on abstract method declarations.

@@

The second condition of \textsc{(T-Intf)} is more complex and is the key to type soundness. Unlike C++ which rejects on ambiguous calls,
\MIM{} rejects on the definition of interfaces when they form a diamond. Consider the case when the second condition is broken: $\mbody(m, J, J)$
is defined but $\mbody(m, I, J)$ is undefined for some $J$ and $m$. This indicates that $m$ is available and unambiguous from the perspective of $J$,
but is ambiguous to $I$ on branch $J$. It means that there are multiple overriding paths of $m$ from $J$ to $I$, which forms a diamond. Hence rejecting
that case meets our expectation. Below is an example (Figure~\ref{fig:examplesmbody} (e)) that illustrates the reason why this condition is needed:
\bruno{what is the purpose of this example:
  state-it upfront please. Is this example meant to ilustrate T-Inf?
  Then it's better to have the example together with the text
  explaining T-inf.} \yanlin{revised.please check whether you're happy with it.}
\vspace{3pt}\begin{lstlisting}
interface T                 { T m() override T { return new T(); } }
interface A extends T       { T m() override T { return new A(); } }
interface B extends T       { T m() override T { return new B(); } }
interface C extends A, B {}

@@

	\framebox{$ \judgeewf \Gamma {e:I} $} \hspace{.5in}
	\tvar \\
	\tinvk \\
	% \tpathinvk \\
	% \tsuperinvk \\
	% \tstaticinvk  \\
	\tnew \\
	\tmethod \\
	\tabsmethod \\
	\tintf \\
	\tanno
\end{mathpar}
\saveSpaceFig
\caption{Typing and subtyping of \name{}.}
\label{fig:typingrules}
\end{figure*}

\subsection{Small-step Semantics and Congruence}
Figure~\ref{fig:smallstep} defines the small-step semantic and
congruence rules of \MIM{}. When evaluating an expression, they
are invoked recursively and alternately, and produce a single value in
the end. \haoyuan{we need to be consistent on paragraph upper/lower case.}

\paragraph{Semantic Rules} \textsc{(S-Invk)} is the only computation rule we need for method invocation.
As a small-step rule and by congruence, it assumes that the receiver and the arguments are already values.
Specifically, the receiver $(J)\new{I}$ indicates the dynamic type $I$
together with the static type $J$. Therefore $\mbody(m, I, J)$ carries out hierarchical dispatching, acquires
the types, the return expression $e_0$ and the interface $I_0$ which provides the most specific method.
Here we use $e_0$ to imply that the return expression is forced to be non-empty because it requires a concrete implementation. Now the
rule reduces method invocation to $e_0$ with substitution.
Parameters are substituted with arguments, and the \lstinline|this| reference is substituted by the receiver,
and in the meanwhile implicit casts are applied, so that static types are recorded via annotations. Finally
there is again an annotation from the return type.

\paragraph{Congruence Rules} \textsc{(C-Receiver)}, \textsc{(C-Args)} and \textsc{(C-FReduce)} are natural congruence rules
on receivers, arguments, and cast-expressions, respectively. \textsc{(C-StaticType)} automatically adds an annotation $I$ to the new
object $\new{I}$. \textsc{(C-AnnoReduce)} merges nested upcasts into a single upcast with the outermost type.

>>>>

\paragraph{Methods}
Original methods and hierarchically overriding methods share the same syntax in our model for simplicity.
The concrete method declaration $\method{I}{m}{I_x}{x}{J}{e}$ introduces a
method named $m$ with result type $I$, parameters $\overline{x}$ of
type $\overline{I_x}$ and the overriding target $J$. The body of the
method simply includes the returned expression $e$. Notably, we have introduced the
\kwoverride{} keyword for two cases: if the overridden interface is exactly the enclosing
interface itself, then such a method is seen as originally defined; otherwise it is a hierarchical overriding method. The definition
of abstract methods is written as $\absmethod{I}{m}{I_x}{x}{J}$, which is
similar to a concrete method but without the method body. 
For simplicity, overloading is not modelled for methods, which
implies that we can uniquely identify a method by its name.

\paragraph{Expressions \& Values}
Expressions can be standard constructs such as variables, method
invocation, object creation, together with cast expressions. 
Object creation is represented by $\new I$\footnote{In Java the corresponding syntax is $\new I\{\}$.}. Fields and primitive types are not modelled in \MIM{}. 
The casts are merely safe upcasts, and in fact, they can be viewed as
annotated expressions, where the annotation indicates its static type.
The coexistence of static and dynamic types is the key to hierarchical dispatch.
A value
``$(I)\new{J}$''
is the final result of multiple reduction steps evaluating an
expression.

@@

\textsc{(T-Invk)} is the typing rule for method invocation.
Naturally, the receiver and the arguments are required to be well-typed.
$\mbody$ is our key function for method lookup that implements the
hierarchical dispatching algorithm. The formal definition will be introduced in Section~\ref{sec:auxdefs}.
Here $\mbody(m, I_0, I_0)$ finds the most specific $m$ above $I_0$. ``Above $I_0$'' specifies
the search space: it not only includes the body of $I_0$, but also its super types.
For the general case, however, the hierarchical invocation $\mbody(m, I, J)$ finds ``the most specific $m$
above $I$ and along path/branch $J$''.\haoyuan{Make sure we mentioned this in Overview.}

In \textsc{(T-Invk)}, as the compilation should not be aware
of the dynamic type, it only requires that invoking $m$ is valid for the static type of the
receiver. The result of $\mbody$ contains the interface that provides the most specific implementation,
the parameters and the return type. We use underscore for the return expression, to imply that an empty return expression
from an abstract method is also acceptable.
% \textsc{(T-PathInvk)} is the typing judgement for a path invocation. Besides the conditions of \textsc{(T-Invk)}, \textsc{(T-PathInvk)} requires the type of receiver to be the subtype of the specified path type. 
% and \textsc{(T-SuperInvk)} checks if the enclosing type directly extends the specified super type.

@@

\textsc{(T-Method)} is more interesting since a method can either be an original method or a hierarchical overriding, though
they share the same syntax and method typing rule. $\mostSpecific(m, I, J)$ is a fundamental function,
used to find ``the most specific interfaces that are above $I$ and
along path $J$, and originally defines $m$'' (see
Section~\ref{sec:auxdefs} for full definition).
By ``most specific interfaces'',
it implies that the inherited super types are excluded. Thus the condition $\mostSpecific(m, I, J) = \{J\}$ indicates a characteristic of a hierarchical overriding: it must override an original method; the overriding is direct and there does not exist any other original method $m$ in between.
Then $\mbody(m, J, J)$ provides the type of the original method, so hierarchical overriding has to preserve the type. Finally the return expression
is type-checked to be subtype of the declared return type. For the definition of an original method, $I$ equals $J$ and the rule is straightforward. \textsc{(T-AbsMethod)} is a similar rule but works on abstract method declarations.

@@

The second condition of \textsc{(T-Intf)} is more complex and is the key to type soundness. Unlike C++ which rejects on ambiguous calls,
\MIM{} rejects on the definition of interfaces when they form a diamond. Consider the case when the second condition is broken: $\mbody(m, J, J)$
is defined but $\mbody(m, I, J)$ is undefined for some $J$ and $m$. This indicates that $m$ is available and unambiguous from the perspective of $J$,
but is ambiguous to $I$ on branch $J$. It means that there are multiple overriding paths of $m$ from $J$ to $I$, which forms a diamond. Hence rejecting
that case meets our expectation. Below is an example (Figure~\ref{fig:examplesmbody} (e)) that illustrates the reason why this condition is needed:
%\bruno{what is the purpose of this example:
%  state-it upfront please. Is this example meant to ilustrate T-Inf?
%  Then it's better to have the example together with the text
%  explaining T-inf.} \yanlin{revised.please check whether you're happy with it.}
\vspace{3pt}\begin{lstlisting}
interface T                 { T m() override T { return new T(); } }
interface A extends T       { T m() override T { return new A(); } }
interface B extends T       { T m() override T { return new B(); } }
interface C extends A, B {}

@@

	\framebox{$ \judgeewf \Gamma {e:I} $} \hspace{.5in}
	\tvar \\
	\tinvk \\
	% \tpathinvk \\
	% \tsuperinvk \\
	% \tstaticinvk  \\
	\tnew \\
	\tanno \\
	\tmethod \\
	\tabsmethod \\
	\tintf
\end{mathpar}
\saveSpaceFig
\caption{Typing and subtyping of \name{}.}
\label{fig:typingrules}
\end{figure*}

\subsection{Small-step Semantics and Congruence}
Figure~\ref{fig:smallstep} defines the small-step semantic and
congruence rules of \MIM{}. When evaluating an expression, they
are invoked recursively and alternately, and produce a single value in
the end. %\haoyuan{we need to be consistent on paragraph upper/lower case.}

\paragraph{Semantic Rules} \textsc{(S-Invk)} is the only computation rule we need for method invocation.
As a small-step rule and by congruence, it assumes that the receiver and the arguments are already values.
Specifically, the receiver $(J)\new{I}$ indicates the dynamic type $I$
together with the static type $J$. Therefore $\mbody(m, I, J)$ carries out hierarchical dispatching, acquires
the types, the return expression $e_0$ and the interface $I_0$ which provides the most specific method.
Here we use $e_0$ to imply that the return expression is forced to be non-empty because it requires a concrete implementation. Now the
rule reduces method invocation to $e_0$ with substitution.
Parameters are substituted with arguments, and the \lstinline|this| reference is substituted by the receiver,
and in the meanwhile the static types are recorded via annotations. Finally the return type $I_e$ is put in the front as an annotation.
\paragraph{Congruence Rules} \textsc{(C-Receiver)}, \textsc{(C-Args)} and \textsc{(C-FReduce)} are natural congruence rules
on receivers, arguments, and cast-expressions, respectively. \textsc{(C-StaticType)} automatically adds an annotation $I$ to the new
object $\new{I}$. \textsc{(C-AnnoReduce)} merges nested upcasts into a single upcast with the outermost type.

### 2017-10-18 NewVersion/sections/Overview.tex
Fix.

\paragraph{\MIM{}'s solution:} To solve this problem it is important to preserve both conflicting methods
during inheritance instead of merging them into a single
method. Therefore \MIM{} accepts the definition of
\lstinline|DrawableDeck|. To disambiguate method calls, 
we can use \emph{up-casts} in \MIM{} to specify the ``branch'' in the
inheritance hierarchy that should be called. The following code
illustrates the use of up-casts for disambiguation:

@@

This example illustrates the basic form of triangle inheritance, where two unintentionally conflicting methods
are accepted by multiple inheritance. Note that C++ supports this feature and also addresses the
ambiguity by up-casts. The code for the above example in C++ is similar.

@@

A special rule for hierarchical overriding is: it can only refine \textbf{original} methods, and cannot jump over original methods with the same signature. For instance, writing \lstinline|"void draw()| \lstinline|override Deck {...}"| is disallowed in \lstinline|DrawableSafeDeck|, because existing two branches are \lstinline|Drawable.draw| and \lstinline|SafeDeck.draw|, while \lstinline|Deck.draw| is already covered. It does not really make sense to refine an old branch. \haoyuan{I think we are abusing font styles. TODO: put all bold text in italic, except for code. Keep FHJ as bold.}

>>>>

\paragraph{\MIM{}'s solution:} To solve this problem it is important to preserve both conflicting methods
during inheritance instead of merging them into a single
method. Therefore \MIM{} accepts the definition of
\lstinline|DrawableDeck|. To disambiguate method calls, 
we can use \emph{upcasts} in \MIM{} to specify the ``branch'' in the
inheritance hierarchy that should be called. The following code
illustrates the use of upcasts for disambiguation:

@@

This example illustrates the basic form of triangle inheritance, where two unintentionally conflicting methods
are accepted by multiple inheritance. Note that C++ supports this feature and also addresses the
ambiguity by upcasts. The code for the above example in C++ is similar.

@@

A special rule for hierarchical overriding is: it can only refine \textit{original} methods, and cannot jump over original methods with the same signature. For instance, writing \lstinline|"void draw()| \lstinline|override Deck {...}"| is disallowed in \lstinline|DrawableSafeDeck|, because existing two branches are \lstinline|Drawable.draw| and \lstinline|SafeDeck.draw|, while \lstinline|Deck.draw| is already covered. It does not really make sense to refine an old branch.
